---
title: "Analysing the BABS data"
author: "Georgie Knight"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document: 
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Dplyr and tidyr
Load the packages and the data. We'll now use the _readr_ package so that we don't have to convert the dates and times.
```{r}
library("dplyr")
library("tidyr")
library("lubridate")
library("readr")
library("ggplot2")
library("ggmap")
trip_read    <- read_csv("trip_full_updated.csv")
status_read  <- read_csv("status_full_updated.csv")
trip         <- dplyr::tbl_df(trip_read)
status       <- dplyr::tbl_df(status_read)

glimpse(trip)
glimpse(status)
```


So now we have our data ready we can play with it.

```{r}

```






```{r}
trip %>% select(Duration) %>% summary()
```

We see that the average trip length is $1046/60 = 17.4$ minutes, the minimum trip length was $1$ minute and the max was nearly $200$ days. That may be a mistake...
Does this vary by location?


```{r}
distinct(trip, landmark)

trip %>% filter(landmark == "San Francisco") %>% 
        select(Duration) %>% 
        summary()
trip %>% filter(landmark == "Mountain View") %>% 
        select(Duration) %>% 
        summary()

trip %>% filter(landmark == "San Jose") %>% 
        select(Duration) %>% 
        summary()

trip %>% filter(landmark == "Palo Alto") %>% 
        select(Duration) %>% 
        summary()

trip %>% filter(landmark == "Redwood City") %>% 
        select(Duration) %>% 
        summary()
```


We can find the most used starting points
```{r}
Count_starting_point <- trip %>% 
  count(Start.Terminal) %>%  
  arrange(desc(n))

Terminal_landmark <- trip %>% 
  select(Start.Terminal, landmark,  start_lat, start_long) %>% 
  distinct( Start.Terminal, .keep_all = TRUE)

Count_starting_point <-
left_join(Count_starting_point , Terminal_landmark, by= c("Start.Terminal"= "Start.Terminal"))

ggplot(Count_starting_point, aes(x = reorder(Start.Terminal, -n), y = n,  
                                 col = as.factor(landmark))) + 
  geom_point()+ 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  labs(col='Area', x = "Start point", y = "Times used") 

San_Fran_map <-qmap(
  location = 'san francisco financial district', 
  source = "osm", 
  zoom =14)

San_Fran_map + 
  geom_point(data =Count_starting_point, aes( y= start_lat, x =start_long, size = n), alpha = 0.5)+
  labs(size = 'Usage')
```

```{r}
San_Fran_trips <-trip %>% 
  filter(landmark == "San Francisco") %>%
  select(Start.Station, End.Station, start_lat, start_long,end_lat, end_long) %>% 
  count(Start.Station, End.Station,start_lat, start_long, end_lat, end_long)
  
San_Fran_map <-qmap(location = 'san francisco financial district', zoom =14)

San_Fran_map + 
  geom_segment(data = San_Fran_trips, aes(y= start_lat, x = start_long, yend =  end_lat, xend = end_long, size = n), alpha = 0.1)+ 
  theme(legend.position="none")
  
```

Let's look at all the trips in a given week

```{r}
San_Fran_trip_select <-trip %>% 
  filter(landmark == "San Francisco") %>%
  filter(between(Date, as.Date("2015-06-01"), as.Date("2015-06-30"))) %>% 
  select(Start.Station, End.Station, start_lat, start_long,end_lat, end_long) %>% 
  count(Start.Station, End.Station,start_lat, start_long, end_lat, end_long)


San_Fran_map + 
  geom_segment(data = San_Fran_trip_select, aes(y= start_lat, x = start_long, yend =  end_lat, xend = end_long, size = n), alpha = 0.1)+ 
  theme(legend.position="none")
```

Or on a particular day

```{r}
San_Fran_trip_select <-trip %>% 
  filter(landmark == "San Francisco") %>%
  filter(Date == as.Date("2015-06-02")) %>% 
  select(Start.Station, End.Station, start_lat, start_long,end_lat, end_long) %>% 
  count(Start.Station, End.Station,start_lat, start_long, end_lat, end_long)

San_Fran_map + 
  geom_segment(data = San_Fran_trip_select, aes(y= start_lat, x = start_long, yend =  end_lat, xend = end_long, size = n), alpha = 0.1)+ 
  theme(legend.position="none")

```



Let's take a look at how bikes available varies in some select stations in San Francisco in a particular week.

```{r}
status_sf_select <- status %>% 
  filter(landmark == "San Francisco") %>% 
  filter(between(station_id, 50, 60)) %>% 
  filter(between(Date, as.Date("2015-06-01"), as.Date("2015-06-08")))
  
ggplot(status_sf_select, aes(x= time, y = bikes_available, 
                   col = as.factor(station_id)))+ 
  geom_line()+
  geom_smooth()+
  labs(col = 'station id', x = 'time', y = ' number of bikes available' )
```


Or on a particular day

```{r}
status_sf_select <- status %>% 
  filter(landmark == "San Francisco") %>% 
  filter(between(station_id, 50, 60)) %>% 
  filter(Date == as.Date("2015-06-01"))
  
ggplot(status_sf_select, aes(x= time, y = bikes_available, 
                   col = as.factor(station_id)))+ 
  geom_line()+
  geom_point()+
  labs(col = 'station id', x = 'time', y = ' number of bikes available' )
```


We'll convert the date to a time column within the lubridate package and create a weekday column too. This will help with analysis.

```{r}
trip <- trip  %>% mutate(
    Date = parse_date_time(Date,"%Y/%m/%d"),
    Weekday = wday(Date, label=TRUE, abbr=FALSE)
                         )
```