---
title: "Analyse_3: Graph analysis."
author: "Georgie Knight"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE, message=FALSE)
```
```{r}
library("dplyr")
library("tidyr")
library("lubridate")
library("readr")
library("ggplot2")
library("ggmap")
library("igraph")
library("popgraph")
library("zipcode")
trip_read    <- read_csv("trip_full_updated2.csv")
status_read  <- read_csv("status_full_updated.csv")
trip         <- dplyr::tbl_df(trip_read)
status       <- dplyr::tbl_df(status_read)
```


## Introduction
We're going to do use some graph theory to analyse the BABS bike network. This will help us figure out which stations are the most integral to the network.

## Create the network
We'll first create a network out of our bike network with stations as the nodes and number of trips between stations as the weights on the edges. We'll normalise these weights.


```{r}
station <-read_csv("201508_station_data.csv")
station <- dplyr::tbl_df(station)
station$name[36] = "Washington at Kearny" #correct the misspellings!
station$name[37] = "Post at Kearny"

tripNumbers <- trip %>%  
  group_by(Start.Station,End.Station) %>% 
  summarise(weight = n())  %>%
  ungroup() %>% 
  mutate(weight = (weight - min(weight))/(max(weight)-min(weight)))

bikeGraph <- graph_from_data_frame(tripNumbers, directed=TRUE) %>% 
set_vertex_attr("latitude", value = station$lat) %>% 
set_vertex_attr("longitude", value = station$long) %>% 
set_vertex_attr("city", value = station$landmark)

```


## Visualise the network
We can firstly visualise our network:

```{r}
location <- c( mean(V(bikeGraph)$longitude), mean(V(bikeGraph)$latitude))
BABSmap <- get_map(location,maptype="roadmap", zoom=10)

bayTrips <-trip %>% 
  select(Start.Station, End.Station, start_lat, start_long,end_lat, end_long) %>% 
  count(Start.Station, End.Station,start_lat, start_long, end_lat, end_long)

ggmap(BABSmap)+
  geom_segment(data = bayTrips, 
               aes(y= start_lat, 
                   x = start_long, 
                   yend =  end_lat, 
                   xend = end_long), alpha = 0.1)+ 
  theme(legend.position="none")+
  geom_nodeset( aes(x=longitude, y=latitude), bikeGraph, size=1, alpha =0.9, col="red")
```

That's a pretty long trip from San Jose to San Francisco!

```{r}
longTrip <- trip %>% 
  filter(startLandmark =="San Jose", endLandmark =="San Francisco") %>% 
  select(Trip.ID, 
         Duration, 
         Start_trip, 
         Start.Station, 
         End_trip, 
         End.Station,
         Subscriber.Type)
longTrip
```

## Degree

The degree of a node tells us how many stations it connects to. We'll calculate the sum of the in and out degrees for each station

```{r}
deg <- degree(bikeGraph, mode= "total")
deg <- sort(deg)
degDF <- data.frame(station = names(deg), degree = deg)
degDF <- dplyr::tbl_df(degDF)
degreeDF <- left_join(degDF, station, by = c("station" = "name"))

ggplot(degreeDF, aes(x=station, y = degree, col = landmark)) +
  geom_point() +
  theme(text = element_text(size=6), axis.text.x = element_text(angle=90, vjust=1))+
  labs(x="Station", y= "Degree", col = "City", title = "BABS station degrees") +
  scale_x_discrete(limits = degreeDF$station)
```

This is interesting but we can use a more discriminatory property. 

## Eigenvalue centrality

We'll now split the network up and analyse the individual cities by calculating the eigenvalue centrality.

```{r}
tripNumbers <- trip %>%  
  filter(startLandmark == "Mountain View", endLandmark == "Mountain View") %>% 
  group_by(Start.Station,End.Station) %>% 
  summarise(weight = n())  %>%
  ungroup() %>% 
  mutate(weight = (weight - min(weight))/(max(weight)-min(weight)))

bikeGraph <- graph_from_data_frame(tripNumbers, directed=TRUE)
bikeE <- eigen_centrality(bikeGraph, directed = TRUE)
bikeEigen <- data.frame(station = names(bikeE$vector), eig = bikeE$vector)
bikeEigen <- dplyr::tbl_df(bikeEigen)
bikeEigen <- mutate(bikeEigen, station = as.character(station))
bikeEigen <- arrange(bikeEigen, eig)
ggplot(bikeEigen, aes(x=station, y = eig))+
  geom_point()+
  theme(text = element_text(size=8), axis.text.x = element_text(angle=90, vjust=1))+
  labs(x="Station", y= "Centrality measure", title = "Mountain View")+
  scale_x_discrete(limits = bikeEigen$station)
```


```{r}
tripNumbers <- trip %>%  
  filter(startLandmark == "Palo Alto", endLandmark == "Palo Alto") %>% 
  group_by(Start.Station,End.Station) %>% 
  summarise(weight = n())  %>%
  ungroup() %>% 
  mutate(weight = (weight - min(weight))/(max(weight)-min(weight)))

bikeGraph <- graph_from_data_frame(tripNumbers, directed=TRUE)
bikeE <- eigen_centrality(bikeGraph, directed = TRUE)
bikeEigen <- data.frame(station = names(bikeE$vector), eig = bikeE$vector)
bikeEigen <- dplyr::tbl_df(bikeEigen)
bikeEigen <- mutate(bikeEigen, station = as.character(station))
bikeEigen <- arrange(bikeEigen, eig)
ggplot(bikeEigen, aes(x=station, y = eig))+
  geom_point()+
  theme(text = element_text(size=8), axis.text.x = element_text(angle=90, vjust=1))+
  labs(x="Station", y= "Centrality measure", title = "Palo Alto")+
  scale_x_discrete(limits = bikeEigen$station)
```


```{r}
tripNumbers <- trip %>%  
  filter(startLandmark == "Redwood City", endLandmark == "Redwood City") %>% 
  group_by(Start.Station,End.Station) %>% 
  summarise(weight = n())  %>%
  ungroup() %>% 
  mutate(weight = (weight - min(weight))/(max(weight)-min(weight)))

bikeGraph <- graph_from_data_frame(tripNumbers, directed=TRUE)
bikeE <- eigen_centrality(bikeGraph, directed = TRUE)
bikeEigen <- data.frame(station = names(bikeE$vector), eig = bikeE$vector)
bikeEigen <- dplyr::tbl_df(bikeEigen)
bikeEigen <- mutate(bikeEigen, station = as.character(station))
bikeEigen <- arrange(bikeEigen, eig)
ggplot(bikeEigen, aes(x=station, y = eig))+
  geom_point()+
  theme(text = element_text(size=8), axis.text.x = element_text(angle=90, vjust=1))+
  labs(x="Station", y= "Centrality measure", title = "Redwood City")+
  scale_x_discrete(limits = bikeEigen$station)
```

```{r}
tripNumbers <- trip %>%  
  filter(startLandmark == "San Francisco", endLandmark == "San Francisco") %>% 
  group_by(Start.Station,End.Station) %>% 
  summarise(weight = n())  %>%
  ungroup() %>% 
  mutate(weight = (weight - min(weight))/(max(weight)-min(weight)))

bikeGraph <- graph_from_data_frame(tripNumbers, directed=TRUE)
bikeE <- eigen_centrality(bikeGraph, directed = TRUE)
bikeEigen <- data.frame(station = names(bikeE$vector), eig = bikeE$vector)
bikeEigen <- dplyr::tbl_df(bikeEigen)
bikeEigen <- mutate(bikeEigen, station = as.character(station))
bikeEigen <- arrange(bikeEigen, eig)
ggplot(bikeEigen, aes(x=station, y = eig))+
  geom_point()+
  theme(text = element_text(size=8), axis.text.x = element_text(angle=90, vjust=1))+
  labs(x="Station", y= "Centrality measure", title = "San Francisco")+
  scale_x_discrete(limits = bikeEigen$station)
```


```{r}
tripNumbers <- trip %>%  
  filter(startLandmark == "San Jose", endLandmark == "San Jose") %>% 
  group_by(Start.Station,End.Station) %>% 
  summarise(weight = n())  %>%
  ungroup() %>% 
  mutate(weight = (weight - min(weight))/(max(weight)-min(weight)))

bikeGraph <- graph_from_data_frame(tripNumbers, directed=TRUE)
bikeE <- eigen_centrality(bikeGraph, directed = TRUE)
bikeEigen <- data.frame(station = names(bikeE$vector), eig = bikeE$vector)
bikeEigen <- dplyr::tbl_df(bikeEigen)
bikeEigen <- mutate(bikeEigen, station = as.character(station))
bikeEigen <- arrange(bikeEigen, eig)
ggplot(bikeEigen, aes(x=station, y = eig))+
  geom_point()+
  theme(text = element_text(size=8), axis.text.x = element_text(angle=90, vjust=1))+
  labs(x="Station", y= "Centrality measure", title = "San Jose")+
  scale_x_discrete(limits = bikeEigen$station)
```

Interestingly we see that the busiest station in each city is located by a train station. This is important if we want to expand into new cities as people seem to be using them as part of their commute. We can investigate this further by analysing the types of trips being taken.

```{r}
 tripTypes <- trip  %>%  
  group_by(Start.Station, End.Station) %>% 
  summarise(count =n()) %>% 
  ungroup() %>% 
  arrange(desc(count)) %>% 
  left_join(station, by = c("Start.Station" = "name"))

```

We can look at the most common trips in each city

```{r}
tripTypes  %>% 
  filter(landmark == "Mountain View")  %>% 
  select(Start.Station, End.Station, count)  %>% 
  mutate(Start.Station = abbreviate(Start.Station, 30), 
         End.Station = abbreviate(End.Station, 30))%>% 
  slice(1:10)

tripTypes  %>% 
  filter(landmark == "Palo Alto")  %>% 
  select(Start.Station, End.Station, count)  %>% 
  mutate(Start.Station = abbreviate(Start.Station, 35), 
         End.Station = abbreviate(End.Station, 35))%>% 
  slice(1:10)

tripTypes  %>% 
  filter(landmark == "Redwood City")  %>% 
  select(Start.Station, End.Station, count)  %>% 
  mutate(Start.Station = abbreviate(Start.Station, 35), 
         End.Station = abbreviate(End.Station, 35)) %>% 
  slice(1:10)

tripTypes  %>% 
  filter(landmark == "San Francisco")  %>% 
  select(Start.Station, End.Station, count)  %>% 
  mutate(Start.Station = abbreviate(Start.Station, 30), 
         End.Station = abbreviate(End.Station, 30)) %>% 
  slice(1:10)

tripTypes  %>% 
  filter(landmark == "San Jose")  %>% 
  select(Start.Station, End.Station, count)  %>% 
  mutate(Start.Station = abbreviate(Start.Station, 30), 
         End.Station = abbreviate(End.Station, 30)) %>% 
  slice(1:10)

```

From which we see how integral the train stations are to this bike network. Who is taking these trips? Is it subscribers?

```{r}
trip %>% filter(Start.Station == "San Francisco Caltrain 2 (330 Townsend)") %>% 
  group_by(Subscriber.Type) %>% summarise(count = n())
```
We see it is mainly subscribers using this station. What time of day is it being used?

```{r}
trip_per_hour_start <- trip %>% 
  filter(Start.Station == "San Francisco Caltrain 2 (330 Townsend)") %>% 
  mutate(sHour = hour(Start_trip))

trip_per_hour_end <- trip %>% 
  filter(End.Station == "San Francisco Caltrain 2 (330 Townsend)") %>% 
  mutate(eHour = hour(End_trip))

ggplot(trip_per_hour_start, aes(x = sHour))+
geom_bar(col = 'grey', fill = "green", alpha = 0.8, 
         width = 0.5, position = position_nudge(x = 0.25))+
geom_bar(data = trip_per_hour_end, aes(x=eHour), 
         col = 'grey', fill = "red", alpha = 0.8, width = 0.5,
         position = position_nudge(x = 0.75))+
labs(x= "Hour", y="Trips", title = "Start(Green) and End(Red) at SF caltrain 2")+
  scale_x_continuous(breaks =seq(0,24,1))
                                                                                  
```

From which we see it it mainly at the hours of people arriving for work and leaving from work.

Where do they live? we can use the zipcode information to answer this combined with the zipcode data and library package.

```{r}
data("zipcode")
zips <- trip %>% 
  filter(Start.Station == "San Francisco Caltrain 2 (330 Townsend)") %>% 
  group_by(Zip.Code) %>% 
  summarise(count = n()) %>% 
  ungroup() %>% 
  mutate(zip = clean.zipcodes(Zip.Code)) %>% 
  merge(zipcode, by.x='zip', by.y='zip') %>% 
  arrange(desc(count))

head(zips)
```

But note there is one subscriber from as far as Hawaii!
```{r}
filter(zips, state == "HI")
```
( I doubt they are commuting though)

Let's look at the top 100 zipcodes which are all in California.

```{r}
zipsShort <- slice(zips, 1:100)
location <- c( mean(zipsShort$longitude), mean(zipsShort$latitude))
BABSmap <- get_map(location,maptype="roadmap", zoom=8)
ggmap(BABSmap)+
  geom_point(data = zipsShort, aes(x=longitude, y=latitude, size = count), alpha = 0.5)+
  labs(x = "longitude", y = "latitude", size = "Count",
       title = "Zip codes of users of SF Caltrain 2 station.")
```
People travel from afar to use this station. Let's look more specifically by choosing the top 20 zipcodes.


```{r}
zipsShorter <- slice(zips, 1:20)
location <- c( mean(zipsShorter$longitude), mean(zipsShorter$latitude))
BABSmap <- get_map(location,maptype="roadmap", zoom=10)
ggmap(BABSmap)+
  geom_point(data = zipsShorter, aes(x=longitude, y=latitude, size = count), alpha = 0.5)+
  labs(x = "longitude", y = "latitude", size = "Count", 
       title = "Zip codes of users of SF Caltrain 2 station.")
```













##Clusters

Let's try to find communities within the city of San Francisco using the infomap algorithm

Here is our network:
```{r}
tripNumbers <- trip %>%  
  filter(startLandmark == "San Francisco", endLandmark == "San Francisco") %>% 
  group_by(Start.Station,End.Station) %>% 
  summarise(weight = n())  %>%
  ungroup() %>% 
  mutate(weight = (weight - min(weight)+0.0000001)/(max(weight)-min(weight)))

stationSF <- station  %>% filter(landmark == "San Francisco") %>% arrange(name)

bikeGraph <- graph_from_data_frame(tripNumbers, directed=TRUE) 
location <- c( mean(stationSF$long), mean(stationSF$lat))
BABSmap <- get_map(location,maptype="roadmap", zoom=14)
```

Let's look at some clustering algorithms
```{r}
clus_eb <- cluster_edge_betweenness(bikeGraph)
clus_op <- cluster_optimal(bikeGraph)
clus_sg <- cluster_spinglass(bikeGraph)
clus_wt <- cluster_walktrap(bikeGraph)

stationSF$eb <-clus_eb$membership
stationSF$op <-clus_op$membership
stationSF$sg <-clus_sg$membership
stationSF$wt <-clus_wt$membership
```

and see what results they give us
```{r}
ggmap(BABSmap)+
  geom_point(data = stationSF, 
             aes( y= lat, 
                  x =long, 
                  col = as.factor(eb)))+
  labs(title = "Edge betweeness", col ="Group")

ggmap(BABSmap)+
  geom_point(data = stationSF, 
             aes( y= lat, 
                  x =long, 
                  col = as.factor(op)))+
  labs(title = "Optimal", col ="Group")

ggmap(BABSmap)+
  geom_point(data = stationSF, 
             aes( y= lat, 
                  x =long, 
                  col = as.factor(sg)))+
  labs(title = "Spin glass", col ="Group")

ggmap(BABSmap)+
  geom_point(data = stationSF, 
             aes( y= lat, 
                  x =long, 
                  col = as.factor(wt)))+
  labs(title = "Walk trap", col ="Group")
```


The useful clusterings seem to be mainly correlated with location, indicating lots of short trips within these clusters.



