---
title: "Combining the  data"
author: "Georgie Knight"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
We'll now combine our data file on station data with the station information

## Dplyr and tidyr

Load the _dplyr_ and _tidyr_ packages which will help us wrangle the data:

```{r}
library("dplyr")
library("tidyr")
```
## Loading up the data

Load up our status data and station information:

```{r}
status_read  <- read.csv("status.csv")
station_read <- read.csv("station.csv")
weather_read <- read.csv("weather.csv")

status_df    <- data.frame(status_read)
station_df   <- data.frame(station_read)
weather_df   <- data.frame(weather_read)

status       <- dplyr::tbl_df(status_df)
station      <- dplyr::tbl_df(station_df)
weather      <- dplyr::tbl_df(weather_df)

status       <- select(status, -X)
station      <- select(station, -X)
weather      <- select(weather, -X)

status <- left_join(status, station)

glimpse(status)
```

The dock count column is now redundant

```{r}
status <- status %>%
  select(-dockcount) %>%
  mutate(installation = as.Date(installation))
glimpse(status)
```

Take a look at some random rows:

```{r]}
randomRows = sample(1:length(status$time), 10, replace=T)
slice(status, randomRows)
```

we'll now create a Date column and convert the landmark to character

```{r}

status <- status %>%
  mutate(Date = as.Date(time)) %>%
  mutate(landmark = as.character(landmark))

weather <- mutate(weather, Date = as.Date(Date))
```

We're now ready to add the weather data

```{r}

status <- left_join(status, weather)
write.csv(status, file="status_full.csv")

```

Let's make a quick visual
```{r}
library(ggplot2)
library(lubridate)
status <- mutate(status, time = ymd_hms(time))
ggplot(status, aes(x= time, y = bikes_available, 
                   col = as.factor(station_id)))+
                                   geom_smooth()
```


