---
title: "Preparing the station data"
author: "Georgie Knight"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
We'll now prepare the station data to be combined with the status data.

## Dplyr and tidyr

Load the _dplyr_ and _tidyr_ packages which will help us wrangle the data:

```{r}
library("dplyr")
library("tidyr")
```



We note the following important information about the stations:

_Although stations were installed prior to 8/29/13 (system launch), no station was active until launch date. Therefore, to accurately capture station popularity, we recommend adjusting all pre-launch installation dates to 8/29/13._

_Station names and locations listed on 201508stationdata.csv represent data that was collected on 8/31/15. However, please note that during 9/1/14 and 8/31/15, 5 stations were moved and 1 station stayed in the same location but changed name._

_Station 23: From 9/1/14 - 10/22/14: This station was located at (37.488501, -122.231061)._

_Station 25: From 9/1/14 - 10/22/14: This station was located at (37.486725, -122.225551). It was previously named "Broadway at Main."_

_Station 49: From 9/1/14 - 2/5/15: This station was located at (37.789625, -122.390264)._ 

_Station 69: From 9/1/14 - 3/11/15: This station was located at (37.776377,-122.39607)._ 

_Station 72: Moved twice. From 9/1/14 - 2/12/15, this station was located at (37.780356, -122.412919). From 2/13/15 to 6/3/15, the station was located at (37.780353, -122.41226)._ 

_Station 80: On 9/1/14, this station changed names from "San Jose Government Center" to "Santa Clara County Civic Center." It did not move._

We'll take the station data and update the installation column.

## Loading up the data

Load up our status data and station information:

```{r}
station_data <- 
  read.csv("C:/Users/Georgie/Desktop/BABS_Data/201508_station_data.csv")
station_df  <- data.frame(station_data)
station  <- dplyr::tbl_df(station_df)
```

We update all the dates prior to system launch to the system launch date. Then convert the data to date format

```{r}
station <-station %>%  
          mutate(
            installation = as.Date(installation, format ="%m/%d/%Y")
                ) %>%
          mutate(
            installation = 
            ifelse(installation < "2013-08-29","2013-08-29",as.character(installation))
                ) %>% 
          mutate(installation =as.Date(installation))
  
glimpse(station)
write.csv(station, file="station.csv")
```




