---
title: "Preparing the trip data"
author: "Georgie Knight"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
We'll now combine our data file on station data with the station information

## Dplyr and tidyr

Load the _dplyr_ and _tidyr_ packages which will help us wrangle the data:

```{r}
library("dplyr")
library("tidyr")
```
## Loading up the data

Load up our status data and station information:

```{r}
trip_read  <- read.csv("201508_trip_data.csv")
trip_df    <- data.frame(trip_read)
trip       <- dplyr::tbl_df(trip_df)
glimpse(trip)
```

We note the following:

_-Trip ID: numeric ID of bike trip_
_-Duration: time of trip in seconds_
_-Start Date: start date of trip with date and time, in PST_
_-Start Station: station name of start station_
_-Start Terminal: numeric reference for start station_
_-End Date: end date of trip with date and time, in PST_
_-End Station: station name for end station_
_-End Terminal: numeric reference for end station_
_-Bike #: ID of bike used_
_-Subscription Type: Subscriber = annual or 30-day member; Customer = 24-hour or 3-day member_
_-Zip Code: Home zip code of subscriber (customers can choose to manually enter zip at kiosk however data is unreliable)_

Combine with the station location data and create a date column


```{r}
station_read  <- read.csv("station.csv")
station_df    <- data.frame(station_read)
station       <- dplyr::tbl_df(station_df)
station_short <-select(station, station_id, lat, long, landmark)

trip <- trip %>% 
  mutate(Date = as.Date(trip$Start.Date[1], format = "%m/%d/%Y")) %>% 
  left_join(station_short, by = c("Start.Terminal" = "station_id")) %>% 
  mutate(start_lat =lat, start_long = long) %>% 
  select(-lat, -long) 

station_short <-select(station, station_id, lat, long)

trip <- trip %>%
  left_join(station_short, by = c("End.Terminal" = "station_id")) %>% 
  mutate(end_lat =lat, end_long = long) %>% 
  select(-lat, -long)

glimpse(trip)
```

Note we have only added the landmark once. We presume that start and end landamrks are the same. 

```{r}
weather_read  <- read.csv("weather.csv")
weather_df    <- data.frame(weather_read)
weather       <- dplyr::tbl_df(weather_df)
weather <- weather %>% 
  select(-X) %>% 
  mutate(Date = as.Date(trip$Start.Date[1], format = "%m/%d/%Y"))
glimpse(weather)

semi_join(trip, weather, by="Date")

glimpse(trip)

write.csv(trip, file="trip_full.csv")

```

