---
title: "Preparing the weather data"
author: "Georgie Knight"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
We'll now prepare the weather data to be combined with the status data

## Dplyr and tidyr

Load the _dplyr_ and _tidyr_ packages which will help us wrangle the data:

```{r}
library("dplyr")
library("tidyr")
```

## Loading up the data

Load up our weather data and station information:

```{r}
weather_data <- read.csv("201508_weather_data.csv")
weather_df   <- data.frame(weather_data)
weather      <- dplyr::tbl_df(weather_df)
glimpse(weather)
```

We note the following information


_Daily weather information per service area, provided from Weather Underground in PST. Weather is listed from north to south (San Francisco, Redwood City, Palo Alto, Mountain View, San Jose)._
_-Precipitation In 	"numeric, in form x.xx but alpha ""T""= trace when amount less than .01 inch"_
_-Cloud Cover 	"scale of 0-8, 0=clear"_	
_-Zip: 94107=San Francisco, 94063=Redwood City, 94301=Palo Alto, 94041=Mountain View, 95113= San Jose"_
_-No data recorded on 8/8/2015 for 94301_


```{r}
weather <- mutate(weather, 
           landmark = ifelse(Zip == 94107, 'San Francisco',
                      ifelse(Zip == 94063, 'Redwood City',
                      ifelse(Zip == 94301, 'Palo Alto',      
                      ifelse(Zip == 94041, 'Mountain View',
                      ifelse(Zip == 95113, 'San Jose',NA       
                             )))))
                 )
```

Take a look at some random rows:

```{r]}
randomRows = sample(1:length(weather$PDT), 10, replace=T)
v <- slice(weather, randomRows)
select(v, PDT, Zip, landmark)
```

Create a Date column
```{r}
weather <- weather %>% 
  mutate(Date = as.Date(PDT, format ="%m/%d/%Y")) %>% 
  select(-PDT)
```

We'll now be able to add the weather data to the status data using the landmark and Date column.

```{r}
write.csv(weather, file="weather.csv")
```
