---
title: "Update the data"
author: "Georgie Knight"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
We'll update our data to take account of the following:

_Station 23: From 9/1/14 - 10/22/14: This station was located at (37.488501, -122.231061)._

_Station 25: From 9/1/14 - 10/22/14: This station was located at (37.486725, -122.225551). It was previously named "Broadway at Main."_

_Station 49: From 9/1/14 - 2/5/15: This station was located at (37.789625, -122.390264)._ 

_Station 69: From 9/1/14 - 3/11/15: This station was located at (37.776377,-122.39607)._ 

_Station 72: Moved twice. From 9/1/14 - 2/12/15, this station was located at (37.780356, -122.412919). From 2/13/15 to 6/3/15, the station was located at (37.780353, -122.41226)._ 

_Station 80: On 9/1/14, this station changed names from "San Jose Government Center" to "Santa Clara County Civic Center." It did not move._

## Dplyr and tidyr

Load the packages and the data:

```{r}
library("dplyr")
library("tidyr")
library("lubridate")
trip_read    <- read.csv("trip_full.csv")
status_read  <- read.csv("status_full.csv")
trip_df      <- data.frame(trip_read)
status_df    <- data.frame(status_read)
trip         <- dplyr::tbl_df(trip_df)
status       <- dplyr::tbl_df(status_df)
```

Firstly the trip data. We'll get the start/end dates/times together into one column using _unite_

```{r}
trip <- trip %>% 
  unite(Start_trip, Start_Date, Start_time, sep= " ", remove = TRUE) %>% 
  unite(End_trip, End_Date, End_time, sep= " ", remove = TRUE)
```

We'll now convert all the relevant data columns into a time format with the correct time zone.

```{r}
trip <- trip %>%
  mutate(Start_trip = ymd_hm(Start_trip, tz ="America/Los_Angeles")) %>% 
  mutate(End_trip   = ymd_hm(End_trip,   tz ="America/Los_Angeles")) %>% 
  mutate(Date =as.Date(Date))
```

Now update the latitudes and longitudes for the start
```{r}
trip <- trip %>% 
  mutate(
start_lat=ifelse(between(Date, as.Date("2014-09-01"), as.Date("2014-10-22")) 
                 & Start.Terminal ==23,37.488501, start_lat)
        ) %>% 
  mutate(
start_long=ifelse(between(Date, as.Date("2014-09-01"),as.Date("2014-10-22")) 
                  & Start.Terminal ==23,-122.231061, start_long)
        ) %>% 
  mutate(
start_lat=ifelse(between(Date, as.Date("2014-09-01"), as.Date("2014-10-22")) 
                 & Start.Terminal ==25,37.486725, start_lat)
        ) %>% 
  mutate(
start_long=ifelse(between(Date, as.Date("2014-09-01"),as.Date("2014-10-22")) 
                  & Start.Terminal ==25,-122.225551, start_long)
        ) %>%
  mutate(
start_lat=ifelse(between(Date, as.Date("2014-09-01"), as.Date("2015-02-05")) 
                 & Start.Terminal ==49,37.789625, start_lat)
        ) %>% 
  mutate(
start_long=ifelse(between(Date, as.Date("2014-09-01"),as.Date("2015-02-05")) 
                  & Start.Terminal ==49,-122.390264, start_long)
        ) %>% 
  mutate(
start_lat=ifelse(between(Date, as.Date("2014-09-01"), as.Date("2015-03-11")) 
                 & Start.Terminal ==69,37.776377, start_lat)
        ) %>% 
  mutate(
start_long=ifelse(between(Date, as.Date("2014-09-01"),as.Date("2015-03-11")) 
                  & Start.Terminal ==69,-122.39607, start_long)
        ) %>%
  mutate(
start_lat=ifelse(between(Date, as.Date("2014-09-01"), as.Date("2015-02-12")) 
                 & Start.Terminal ==72,37.780356, start_lat)
        ) %>% 
  mutate(
start_long=ifelse(between(Date, as.Date("2014-09-01"),as.Date("2015-02-12")) 
                  & Start.Terminal ==72,-122.412919, start_long)
        ) %>%
  mutate(
start_lat=ifelse(between(Date, as.Date("2015-02-13"), as.Date("2015-06-03")) 
                 & Start.Terminal ==72,37.780353, start_lat)
        ) %>% 
  mutate(
start_long=ifelse(between(Date, as.Date("2015-02-13"),as.Date("2015-06-03")) 
                  & Start.Terminal ==72,-122.41226, start_long)
        )
```        

Let's check that this has worked!


```{r}
check <-select(trip, Date, Start.Terminal, start_lat, start_long)
check <-filter(check, Start.Terminal ==69)
filter(check, Date > as.Date("2015-03-11"))
filter(check, Date < as.Date("2015-03-11"))
```  
  





Now update the latitudes and longitudes for the end
```{r}
trip <- trip %>% 
  mutate(
end_lat=ifelse(between(Date, as.Date("2014-09-01"), as.Date("2014-10-22")) 
               & End.Terminal ==23,37.488501, end_lat)
        ) %>% 
  mutate(
end_long=ifelse(between(Date, as.Date("2014-09-01"),as.Date("2014-10-22")) 
                & End.Terminal ==23,-122.231061, end_long)
        ) %>% 
  mutate(
end_lat=ifelse(between(Date, as.Date("2014-09-01"), as.Date("2014-10-22")) 
               & End.Terminal ==25,37.486725, end_lat)
        ) %>% 
  mutate(
end_long=ifelse(between(Date, as.Date("2014-09-01"),as.Date("2014-10-22")) 
                & End.Terminal ==25,-122.225551, end_long)
        ) %>%
  mutate(
end_lat=ifelse(between(Date, as.Date("2014-09-01"), as.Date("2015-02-05")) 
               & End.Terminal ==49,37.789625, end_lat)
        ) %>% 
  mutate(
end_long=ifelse(between(Date, as.Date("2014-09-01"),as.Date("2015-02-05")) 
                & End.Terminal ==49,-122.390264, end_long)
        ) %>% 
  mutate(
end_lat=ifelse(between(Date, as.Date("2014-09-01"), as.Date("2015-03-11")) 
               & End.Terminal ==69,37.776377, end_lat)
        ) %>% 
  mutate(
end_long=ifelse(between(Date, as.Date("2014-09-01"),as.Date("2015-03-11")) 
                & End.Terminal ==69,-122.39607, end_long)
        ) %>%
  mutate(
end_lat=ifelse(between(Date, as.Date("2014-09-01"), as.Date("2015-02-12")) 
               & End.Terminal ==72,37.780356, end_lat)
        ) %>% 
  mutate(
end_long=ifelse(between(Date, as.Date("2014-09-01"),as.Date("2015-02-12")) 
                & End.Terminal ==72,-122.412919, end_long)
        ) %>%
  mutate(
end_lat=ifelse(between(Date, as.Date("2015-02-13"), as.Date("2015-06-03")) 
               & Start.Terminal ==72,37.780353, end_lat)
        ) %>% 
  mutate(
end_long=ifelse(between(Date, as.Date("2015-02-13"),as.Date("2015-06-03")) 
                & Start.Terminal ==72,-122.41226, end_long)
        )
```  


Now for the status data
```{r}
status <- mutate(status, Date = as.Date(Date))
```

```{r}
status <- status %>% 
  mutate(
lat=ifelse(between(Date, as.Date("2014-09-01"), as.Date("2014-10-22")) 
               & station_id == 23, 37.488501, lat)
        ) %>% 
  mutate(
long=ifelse(between(Date, as.Date("2014-09-01"),as.Date("2014-10-22")) 
               &  station_id == 23, -122.231061, long)
        ) %>% 
  mutate(
lat=ifelse(between(Date, as.Date("2014-09-01"), as.Date("2014-10-22")) 
               & station_id == 25, 37.486725, lat)
        ) %>% 
  mutate(
long=ifelse(between(Date, as.Date("2014-09-01"),as.Date("2014-10-22")) 
               & station_id == 25, -122.225551, long)
        ) %>%
  mutate(
lat=ifelse(between(Date, as.Date("2014-09-01"), as.Date("2015-02-05")) 
               & station_id == 49, 37.789625, lat)
        ) %>% 
  mutate(
long=ifelse(between(Date, as.Date("2014-09-01"),as.Date("2015-02-05")) 
                & station_id == 49, -122.390264, long)
        ) %>% 
  mutate(
lat=ifelse(between(Date, as.Date("2014-09-01"), as.Date("2015-03-11")) 
               & station_id == 69, 37.776377, lat)
        ) %>% 
  mutate(
long=ifelse(between(Date, as.Date("2014-09-01"),as.Date("2015-03-11")) 
                & station_id == 69, -122.39607, long)
        ) %>%
  mutate(
lat=ifelse(between(Date, as.Date("2014-09-01"), as.Date("2015-02-12")) 
               & station_id == 72, 37.780356, lat)
        ) %>% 
  mutate(
long=ifelse(between(Date, as.Date("2014-09-01"),as.Date("2015-02-12")) 
                & station_id == 72, -122.412919, long)
        ) %>%
  mutate(
lat=ifelse(between(Date, as.Date("2015-02-13"), as.Date("2015-06-03")) 
               & station_id == 72, 37.780353, lat)
        ) %>% 
  mutate(
long=ifelse(between(Date, as.Date("2015-02-13"),as.Date("2015-06-03")) 
                & station_id == 72, -122.41226, long)
        )
```



```{r}
write.csv(status, file="status_full_updated.csv")

write.csv(trip, file="trip_full_updated.csv")
```